Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-24 demystify.tex
updates

% \section{How much time did you spend looking at an application?}\label{sec:time-spent}
% Short answer: less than 5 mins for quick rejection; 15+ mins for an acceptance.  On average, about 10 mins.

% The GMU evaluation system allows us to download a single pdf that consists of a summary page (e.g., school info, GPA, testscores), then transcripts, standard test results, LoRs, CV, SOP, writing samples So I typically read the application in this order.  Then I make notes on what I've seen, basically evidence that support my evaluation.

% I first look at the summary for school, GPA, est scores (essentially looking for very low GPA or testscores that do not meet the minimum requirements. Then I skim on transcripts look for low/failing grades on CS and related courses.  For these I make comments such as "unknown international school with good GPA, too many low grades on CS and related courses" or "good GPA from top school XYZ". I do not look at details on tests (the main scores are already given in the first summary page).

% Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experiences to distinguish between a strong LoRs from the rest). Comment examples For LoRs include "very strong LoRs from well-known prof. XYZ talking about research experiences" or "generic LoRs, mostly about course work".  I also note any unusual things in the LoRs.

% Then the SOP, which I read carefully (of course if it is bad then it would a quick read). Then my notes for SOP include things such as "SOP mostly on course work, no clear motivation, generic" , or "very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned about publications then I'll make notes such as "has X papers at top venues" or "has X papers but at unknown places".  I will not go online search for the papers (unless they upload them as sample papers) .

% I also skim over sampling papers, essentially looking over the abstract to see if reads well.  I would make notes such as "have a paper under submission to unknown venue, but reads well".

%At GMU we use an evaluation system that allows us to download a single, large pdf that consists of everything submitted by the applicant, e.g., GPA, transcripts, test scores, CV, LoRs, SOP, sample writings, etc.   The first couple of pages show the summary including school, major, GPA, test scores.

>>>>

\section{How much time did you spend looking at my application?}\label{sec:time-spent}

Short answer: this is highly dependent on the adcom member.  For me, it is likely about 10 minutes on average (less than 5 mins for definite rejection; 15+ mins for an acceptance). Note that my earlier years, I spent more time, but now I have more experience and can quickly tell if an application is good or bad.

The GMU evaluation system allows us to download a single pdf that consists of a summary page (e.g., school info, GPA, standard test scores), then transcripts, test results, LoRs, CV, SOP, writing samples (if any). So I typically scan the application in this order.  I also make notes on what I've seen, basically evidence that support my evaluation.

I first skim the summary for school, GPA, test scores to looking for red flags like low GPA or test scores that do not meet the minimum requirements. Then I move to transcripts to look for low/failing grades on CS and related courses.  For these I make comments such as ``- unknown international school with good GPA, too many low grades on CS and related courses" or ``+ very good GPA from top school XYZ". I do not look at test results as the main scores are already given in the summary page.

Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experiences to distinguish between a strong LoRs from the rest). My comments for LoRs include ``+ very strong LoRs from well-known prof. XYZ talking about research experiences'' or ``- generic LoRs, mostly about course work''.  I also note any unusual or unique details in the LoRs, e.g., ``+ LoR mentioned about student's significant contribution on well-known open-source project X''. Sometimes a LoR came directly from a GMU professor who mentioned they already know the student and intend to offer RA, which is a very strong signal for admission (and my comment would be ``Prof. X making RA offer'').

I skim quickly over CV, looking for paper publications, research work experiences, and other relevant achievements such as winning prizes in well-known international programming/math competitions.  My notes for CV include things such as ``+ has X papers at top venues'' or ``+ gold medal at X''. Sometimes I also note about unusual or unique experiences, e.g., ``+ has a patent on X''.

Next is the SOP, which I read carefully (of course if it is bad then it would a quick read). My notes for SOP include things such as ``- SOP mostly on course work, no clear motivation, generic" , or ``+ very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned about publications then I would have ``+ has X papers at top venues" or ``+/- has X papers but at unknown places''.  I will not go online search for the papers (unless they upload them as sample papers).  I also skim over sampling papers, essentially looking over the abstract to see if reads well, and make notes such as ``+ have a paper under submission to unknown venue, but reads well". If the student mentioned about a project, I might also look at it (e.g., if it is on Github) and make notes such as ``+ has a project on Github with X stars''.

At this point I have enough to make up my decision, which I enter to the evaluation system. We have several options for decision, but typically I just use \emph{reject} and \emph{admit with full funding}.  For the latter, I also mention funding sources (e.g., RA from prof. X or TA). I can also recommend very strong students for the University Presidential fellowship.  Finally, the system has text box where I put my comments to support my decision.
